\thetitle

### 2021-03-03 writing/challenges.tex
touch ups for consistent use of mutant/mutated module and module/component

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutant component.  A
debugging scenario results from removing the types from the mutant. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About four fifths of the original mutators
survived this first scrutiny testing. All of the remaining ones yield {\em
interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).

>>>>

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutated component.  A
debugging scenario results from removing the types from the mutated component. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About four fifths of the original mutators
survived this first scrutiny testing. All of the remaining ones yield {\em
interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).

### 2021-03-03 writing/discussion.tex
discussion: last edit
touch ups for consistent use of mutant/mutated module and module/component

Relatedly, the experimental setup hides the type choice a rational programmer must make. When the
run-time checks signal an impedance mismatch in the real world, a programmer
does not have a typed component ready to swap in. Instead, the programmer must
come up with the next set of types---and this means the programmer must make
choices. It is normally possible to assign consistent types to variables in a
component in different ways. The creation and curation of the benchmarks over
many years has driven home this lesson but, fortunately, it has also shown that
the types are in most cases reasonably canonical.  The authors therefore conjecture that
a rational programmer would in most cases come up with an equivalent type
assignment in order to extend a trail as the experimental method describes.

@@

The results in section~\ref{sec:results} also show that the cost of Transient
blame is quite high. Under the Transient semantics, some of the debugging
scenarios exceed the 4-minute timeout or the 6GB-memory limit. To put those
limits into context, the fully typed and fully untyped benchmarks all complete
in a few seconds with minimal memory usage. Furthermore, none of the mixed
Natural configurations hit these limits, and with the blame map turned off,
the Transient semantics also runs these programs in a short amount of time and
well within the memory limits. In short, even though the Transient rational
programmer appears to do well in the experiment, the implementation of the
Transient blame strategy might be unrealistic.

>>>>

Relatedly, the experimental setup hides the type choice a rational programmer must make. When the
run-time checks signal an impedance mismatch in the real world, a programmer
does not have a typed module ready to swap in. Instead, the programmer must
come up with the next set of types---and this means the programmer must make
choices. It is normally possible to assign consistent types to variables in a
module in different ways. The creation and curation of the benchmarks over
many years has driven home this lesson but, fortunately, it has also shown that
the types are in most cases reasonably canonical.  The authors therefore conjecture that
a rational programmer would in most cases come up with an equivalent type
assignment in order to extend a trail as the experimental method describes.

@@

The results in section~\ref{sec:results} also show that the cost of Transient
blame is quite high. Under the Transient semantics, some of the debugging
scenarios exceed the 4-minute timeout or the 6GB-memory limit. To put those
limits into context, the fully typed and fully untyped benchmarks all normally complete
in a few seconds with minimal memory usage. Furthermore, none of the mixed
Natural configurations hit these limits, and with the blame map turned off,
the Transient semantics also runs these programs in a short amount of time and
well within the memory limits. In short, even though the Transient rational
programmer appears to do well in the experiment, the implementation of the
Transient blame strategy might be unrealistic.

### 2021-03-03 writing/rational-effort.tex
touch ups for consistent use of mutant/mutated module and module/component

In addition to success and failure information for the various trails, the
experiment records the number of components a rational programmer has
to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
\rvert$). This number can serve as an additional metric to compare modes.

@@

Measuring effort can also reveal whether the observed effectiveness of the
rational programmer is an artifact of pure chance or not.
In particular, the effort distribution for a mode of the rational programmer can be compared with that of
the random mode that ignores error information entirely and instead selects which
component to type randomly.  Since each mutant has a finite number of
components, random mode trails are always successful. However, the random mode's effort
distribution will be thinly spread out across the range of trail lengths possible
in the set of debugging scenarios.
In contrast, the effort distribution of other modes should be quite different if their effectiveness is not coincidental.

>>>>

In addition to success and failure information for the various trails, the
experiment records the number of modules a rational programmer has
to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
\rvert$). This number can serve as an additional metric to compare modes.

@@

Measuring effort can also reveal whether the observed effectiveness of the
rational programmer is an artifact of pure chance or not.
In particular, the effort distribution for a mode of the rational programmer can be compared with that of
the random mode that ignores error information entirely and instead selects which
module to type randomly.  Since each mutant has a finite number of
modules, random mode trails are always successful. However, the random mode's effort
distribution will be thinly spread out across the range of trail lengths possible
in the set of debugging scenarios.
In contrast, the effort distribution of other modes should be quite different if their effectiveness is not coincidental.

### 2021-03-03 writing/rational-natural.tex
touch ups for consistent use of mutant/mutated module and module/component

The Natural semantics assigns blame to exactly one boundary.  A blame assignment
has the following specific meaning: the typed component may make incorrect type
assumptions about the untyped component in its interface, or the correct
interface exposes a bug in the untyped component (or its dependencies). Our setup rules out the first
alternative (but see section~\ref{sec:conclusion}), and therefore the rational
programmer extends the trail to a scenario that swaps out the untyped component
for its typed counterpart.

Formally,
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$, where
$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails because this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenario successfully pinpoints the location of
the bug. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff (the program for) $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that end in a scenario that does not reveal the bug statically, yet also does not blame
an untyped component. Thus the rational programmer gets
no hints from the gradual type system on how to continue the search for the bug.

>>>>

The Natural semantics assigns blame to exactly one boundary.  A blame assignment
has the following specific meaning: the typed module may make incorrect type
assumptions about the untyped module in its interface, or the correct
interface exposes a bug in the untyped module (or its dependencies). Our setup rules out the first
alternative (but see section~\ref{sec:conclusion}), and therefore the rational
programmer extends the trail to a scenario that swaps out the untyped
module
for its typed counterpart.

Formally,
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$, where
$\blame{\system, \conf}$ denotes the module that $\conf$ blames.
\end{quote}

When the buggy untyped module of a program is replaced by the typed counterpart,
the type checker fails because this module causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenario successfully pinpoints the location of
the bug. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff (the program for) $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that end in a scenario that does not reveal the bug statically, yet also does not blame
an untyped module. Thus the rational programmer gets
no hints from the gradual type system on how to continue the search for the bug.

### 2021-03-03 writing/rational-stuff.tex
touch ups for consistent use of mutant/mutated module and module/component

Applying a mutator to module $\component^*$ of $\system$ acts like a homomorphism
on the generated lattice. The two lattices differ only in the mutant of
$\component^*$. Given this formulation, debugging scenarios are those points in
this new lattice that do not contain the mutant; at all other points, the type
checker points out the type-level mistake in the mutant of $\component^*$.

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The starting scenario $\conf_0$ of a
trail---also referred to as the \emph{root} (of the trail)---is the initial
debugging scenario.  Starting from such a root, the rational
programmer steps along a trail. If the (program for this) scenario type-checks,
the rational programmer runs the program until it raises a run-time error.  The rational programmer
uses the information in the error to decide which component to equip with
types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus
creates a trail. The trail's construction ends when it reaches a scenario that
contains the mutant module because the type checker rejects its typed version
outright.  At this point, the source of the impedance mismatch is identified.

>>>>

Applying a mutator to module $\component^*$ of $\system$ acts like a homomorphism
on the generated lattice. The two lattices differ only in the mutated
$\component^*$. Given this formulation, debugging scenarios are those points in
this new lattice that do not contain the mutated module; at all other points, the type
checker points out the type-level mistake in the mutated $\component^*$.

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The starting scenario $\conf_0$ of a
trail---also referred to as the \emph{root} (of the trail)---is the initial
debugging scenario.  Starting from such a root, the rational
programmer steps along a trail. If the (program for this) scenario type-checks,
the rational programmer runs the program until it raises a run-time error.  The rational programmer
uses the information in the error to decide which module to equip with
types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus
creates a trail. The trail's construction ends when it reaches a scenario that
contains the mutated module because the type checker rejects its typed version
outright.  At this point, the source of the impedance mismatch is identified.

### 2021-03-03 writing/rational-transient.tex
touch ups for consistent use of mutant/mutated module and module/component

The Transient semantics assigns blame to a set of components. The blame
assignment says that the value witnessing the impedance mismatch may have
crossed the boundaries between elements in the set, and each crossing
checked the value's type in a shallow manner.

This ambiguity in Transient blame raises the question of how the rational
programmer should react when the language produces a blame set. Our answer
is that the rational programmer has at least two reasonable options. The first
one is to select the component that is added to the blame set first and
assign types to only that one---after all, if fully checked, the types of this
first component should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to select the
component that is added to the blame set last, effectively interpreting the
blame set as a boundary-aware stack.

@@

\begin{quote}
\it A \emph{Transient-first blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
first component Transient adds to the blame set for $\conf$.
\end{quote}

\begin{quote}
\it A \emph{Transient-last blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
  \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
last component Transient adds to the blame set for $\conf$.
\end{quote}

>>>>

The Transient semantics assigns blame to a set of modules. The blame
assignment says that the value witnessing the impedance mismatch may have
crossed the boundaries between elements in the set, and each crossing
checked the value's type in a shallow manner.

This ambiguity in Transient blame raises the question of how the rational
programmer should react when the language produces a blame set. Our answer
is that the rational programmer has at least two reasonable options. The first
one is to select the module that is added to the blame set first and
assign types to only that one---after all, if fully checked, the types of this
first module should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to select the
module that is added to the blame set last, effectively interpreting the
blame set as a boundary-aware stack.

@@

\begin{quote}
\it A \emph{Transient-first blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
first module Transient adds to the blame set for $\conf$.
\end{quote}

\begin{quote}
\it A \emph{Transient-last blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
  \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
last module Transient adds to the blame set for $\conf$.
\end{quote}

### 2021-03-03 writing/results.tex
touch ups for consistent use of mutant/mutated module and module/component

While most blame trail failures follow the above pattern, a few do
not.  Breaking down the failure reasons for Natural blame (1748 in
total) reveals an additional cause. For a small set of debugging scenarios
(40), Natural produces a run time type error blaming a non-buggy
already-typed component. All these cases are due to known open issues with Typed
Racket and class contracts.

@@

There are two immediate take-aways from the figure. First, the effort for
successfully debugging interesting scenarios (in green) for the random
mode of the rational programmer is highly spread out, as
expected. In contrast, in the other modes, successful effort coalesces at
the left side of the plot, meaning that in most cases the programmer needs
to type a single component to debug a scenario.

>>>>

While most blame trail failures follow the above pattern, a few do
not.  Breaking down the failure reasons for Natural blame (1748 in
total) reveals an additional cause. For a small set of debugging scenarios
(40), Natural produces a run time type error blaming a non-buggy
already-typed module. All these cases are due to known open issues with Typed
Racket and class contracts.

@@

There are two immediate take-aways from the figure. First, the effort for
successfully debugging interesting scenarios (in green) for the random
mode of the rational programmer is highly spread out, as
expected. In contrast, in the other modes, successful effort coalesces at
the left side of the plot, meaning that in most cases the programmer needs
to type a single module to debug a scenario.
